---
title: "Reproducible Research - Course Project 1"
author: "Karen Reijneveld"
date: "14 januari 2018"
output: html_document
---
# Loading and preprocessing the data
Download, unzip and read the data:
```{r}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile = "activity.zip")
unzip("activity.zip")
data <- read.csv("activity.csv", skipNul = T, row.names=NULL)

#Transform the date column into a real date: 
library(lubridate)
data$date <- ymd(data$date)
```

#What is the mean total number of steps taken per day?


```{r totalstepsperday}
library(dplyr)
datadays <- data
datadays <- arrange(datadays, date)
datadays <- group_by(datadays, date)
datadays <- summarize(datadays, steps = sum(steps))
datadays <- subset(datadays, complete.cases(datadays))
```

I tried to do this with a pipeline, but I could not get this to work:   
datadays %>% data %>%  
        arrange(date) %>%  
        group_by(date) %>%  
        summarize(steps = sum(steps)) %>%  
        subset(complete.cases(datadays))  
        
## Histogram of the total number of steps taken each day
```{r histogram}
hist(datadays$steps, main = "Total number of steps taken each day", 
     xlab = "number of steps/day", col = "yellow")
```

## Mean and median number of steps taken each day
```{r meanmedian}
options(scipen = 999) #to avoid scientific notation
mean <- round(mean(datadays$steps), digits = 0)
median <- round(median(datadays$steps), digits = 0)
```

The mean is `r mean` and the median is `r median`.  

#What is de average daily activity pattern?

Below, the mean number of steps is calculated for each interval number. 
```{r}
library(dplyr)
datainterval <- data
datainterval <- arrange(datainterval, interval)
datainterval <- group_by(datainterval, interval)
datainterval <- subset(datainterval, complete.cases(datainterval))
datainterval <- summarize(datainterval, steps = mean(steps))
```

##Time series plot of the average number of steps taken

```{r timeseriesplot}
plot(datainterval, type = "l", col = "green")
```

##The 5-minute interval that, on average, contains the maximum number of steps
Interval number 835 contains 206 steps (on average).
```{r}
print(subset(datainterval, steps == max(steps)))
```

#Imputing missing values
The total number of rows with NAs is 2304. 
```{r}
table(complete.cases(data))
```
##Code to describe and show a strategy for imputing missing data

```{r}
names(datainterval) <- c("interval", "meansteps")
dataNew = merge(data, datainterval, by.y = "interval")
dataNew <- mutate(dataNew, imputedsteps = 
                      ifelse(is.na(steps), meansteps, steps))
dataNew <- dataNew[, c(5,3,1)]
names(dataNew) <- c("steps", "date", "interval")
```
##Histogram of the total number of steps taken each day after missing values are imputed

```{r histogram2}
library(dplyr)
datadays <- dataNew
datadays <- arrange(datadays, date)
datadays <- group_by(datadays, date)
datadays <- summarize(datadays, steps = sum(steps))
datadays <- subset(datadays, complete.cases(datadays))

hist(datadays$steps, main = "Total number of steps (imputed) taken each day", 
     xlab = "number of steps/day", col = "purple")
```

```{r ref.label = "meanmedian"}
```

The new mean is `r mean` and the new median is `r median`. Both are now the same as the mean that ignored the missing values. The influence is very small: only 1 step... 

##Are there any differences in activity patterns between weekdays and weekends?
Panel plot comparing the average number of steps taken per 5 minute interval across weekdays and weekends

```{r panelplot}
#create a new variable 'day' 
#with the dutch words for saturday and sunday
dataNew <- mutate(dataNew, day = 
                          ifelse(weekdays(date)=="zaterdag", "weekend",
                          ifelse(weekdays(date)=="zondag","weekend", 
                                 "weekday")))
#convert it to a factor variable
dataNew$day <- as.factor(dataNew$day)

#sort and group the data
dataNew <- arrange(dataNew, interval)
dataNew <- group_by(dataNew, interval)

#calculate mean per interval on weekdays
dataWkdy <- dataNew
dataWkdy <- subset(dataWkdy, day == "weekday")
dataWkdy <- summarize(dataWkdy, steps = mean(steps))
dataWkdy <- mutate(dataWkdy, day = "weekday")

#calculate mean per interval in the weekend
dataWknd <- dataNew
dataWknd <- subset(dataWknd, day == "weekend")
dataWknd <- summarize(dataWknd, steps = mean(steps))
dataWknd <- mutate(dataWknd, day = "weekend")

dataWk <- rbind(dataWkdy, dataWknd)

#make the plots
library(lattice)
xyplot(steps ~ interval | day, dataWk, type = "l", layout = c(1,2))
```
